A Fundamental Measure Theory Functional for Hard-Sphere Contact Densities

### 2012-02-07 papers/contact/paper.tex
various minor contact-related improvements Tested everything on bingley.

  Finally, we get an answer like
  \begin{align}
    \frac{\delta A_{HS}}{\delta R(\mathbf{r})} &=
    \int \mathbf{dr}' \left\{
    \left(
    \frac{n_0(\mathbf{r}')}{1 - n_3(\mathbf{r}')}
    + \frac{n_1n_2 - \mathbf{n}_{V1}\cdot\mathbf{n}_{V2}}{(1 -
      n_3(\mathbf{r}'))^2}
    + \frac{1}{4\pi}\frac{
      \frac13 n_2^3 - n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}
    }{
      (1-n_3(\mathbf{r}'))^3
    }
    \right) n(\mathbf{r}') \delta(|\mathbf{r}-\mathbf{r}'| - R)
    + \cdots
    \right\}
  \end{align}
  This comes out to quite a mess.  It wouldn't be infeasible to compute
  this, but it wouldn't be efficient or pretty either.
}

@@

  So we need to find the small change in free energy that is caused by
  these small changes in the shape and size of the hard spheres that are
  in contact with this point.  In the following section, we will work
  out the changes in the weighted densities at position $\mathbf{r}'$
  when point of contact we are considering is $\mathbf{r}$.

  The packing fraction $n_3$ is changed at the point where the sphere is
  extended, since a few more spheres will overlap this point.
  \begin{align}
    \Delta n_3(\mathbf{r}') &= n(\mathbf{r})\Delta A \Delta R
    \delta(|\mathbf{r}-\mathbf{r}'| - R)
  \end{align}
  The density $n_0$ is the average density, averaged over a spherical
  shell.  As such, shifting the position of a shell just shifts which
  location we are averaging over.
  \begin{align}
    \Delta n_0(\mathbf{r}') &=
    \frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}\cdot
    \mathbf{\nabla}n(\mathbf{r}) \Delta A \Delta R
    \delta(|\mathbf{r}-\mathbf{r}'|-R)
  \end{align}
  The change in $n_2$ comes from two terms: a change in the density value, as
  we saw in $n_0$, and a change in the area of the segment.
  \begin{align}
    \Delta n_2(\mathbf{r}') &=
    \left(n\left(\mathbf{r} + \Delta R
    \frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}\right)\left(1+2\frac{\Delta
      R}{R}\right)- n(\mathbf{r})\right) \Delta A
    \delta(|\mathbf{r}-\mathbf{r}'|-R) \\
    &=
    \left(\frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}\cdot
    \mathbf{\nabla}n(\mathbf{r}) + 2 \frac{n(\mathbf{r})}{R}
    \right) \Delta A \Delta R
    \delta(|\mathbf{r}-\mathbf{r}'|-R)
  \end{align}
  The vector density $\mathbf{n}_{V2}$ is similar to the density
  $n_2$: both changes in the density value and changes in the surface
  area play a role.
  \begin{align}
    \Delta \mathbf{n}_{V2}(\mathbf{r}') &=
    \frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}
    \left(n\left(\mathbf{r} + \Delta R
    \frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}\right)\left(1+2\frac{\Delta
      R}{R}\right)- n(\mathbf{r})\right) \Delta A
    \delta(|\mathbf{r}-\mathbf{r}'|-R) \\
    &=
    \frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}
    \left(\frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}\cdot
    \mathbf{\nabla}n(\mathbf{r}) + 2 \frac{n(\mathbf{r})}{R}
    \right) \Delta A \Delta R
    \delta(|\mathbf{r}-\mathbf{r}'|-R) 
  \end{align}
  The density $n_1$ (and similarly $\mathbf{n}_{V1}$) can be roughly
  interpreted as the density weighted by the area-averaged mean
  curvature~\cite{oversteegen2005general, roth2010review}.  When we
  extend the sphere at the contact point, we will introduce regions of
  both positive and negative mean curvature.  In principle, we can do
  this in such a way that the two portions cancel, and hence we will
  assume that
  \begin{align}
    \Delta n_1(\mathbf{r}') &= 0 \\
    \Delta \mathbf{n}_{V1}(\mathbf{r}') &= 0
  \end{align}
  In fact, the negative and positive portions are spatially separated,
  so these terms should probably include a derivative term, which I
  should work out.  Particularly as it may even happen to cancel a
  different derivative term that we then won't have to drop!

  We need to then find the change in Helmholtz free energy due to this
  change at position $\mathbf{r}$:
  \begin{align}
    \Delta A_{HS} &= \int \mathbf{dr}'
    \frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')}\Delta n_3(\mathbf{r}')
    +
    \frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')}\Delta n_2(\mathbf{r}')
    +
    \frac{\delta A_{HS}}{\delta \mathbf{n}_{V2}(\mathbf{r}')} \cdot \Delta \mathbf{n}_{V2}(\mathbf{r}')
  \end{align}
}

>>>>

  Finally, we get an answer like
  \begin{align}
    \frac{\delta A_{HS}}{\delta R(\mathbf{r})} &=
    \int \mathbf{dr}' \left\{
    \left(
    \frac{n_0(\mathbf{r}')}{1 - n_3(\mathbf{r}')}
    + \frac{n_1n_2 - \mathbf{n}_{V1}\cdot\mathbf{n}_{V2}}{(1 -
      n_3(\mathbf{r}'))^2}
    + \frac{1}{4\pi}\frac{
      \frac13 n_2^3 - n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}
    }{
      (1-n_3(\mathbf{r}'))^3
    }
    \right) n(\mathbf{r}') \delta(|\mathbf{r}-\mathbf{r}'| - R) \right.
    \\
    & \left.
    + \frac{\delta A}{\delta n_2(\rr')} n(\rr) \delta'(|\rr-\rr'|-R) \cdots
    \right\}
  \end{align}
  This comes out to quite a mess.  It isn't be infeasible to compute
  this, but it is challenging.
}

@@
